Subsidiary Recursion in Coq

### 2022-01-15 paper/paper.tex
almost done, but paper is too long
I skimped on the example proofs and still I am a page over without
writing a conclusion or finishing the related work section.  So I will
need to do some trimming once those last things are written.

This paper presents a functor-generic derivation of terminating
subsidiary recursion and induction in Coq.  We should emphasize that
this is a derivation of this recursion scheme within the type theory
of Coq.  No axioms or other modifications to Coq of any kind are
required. Based on this derivation, we present several example
functions like \verb|wordsBy|, and prove theorems about them.  For
example, we prove the expected property that the sublists returned by
\verb|wordsBy| consist of elements satisfying \verb|not . p|.
For another, we give a definition of run-length encoding as a
subsidiary recursion using \verb|span|, and prove that encoding and
then decoding returns the original list.  Our approach applies to the
standard datatypes in the Coq library, and does not require switching
libraries or datatype definitions.

@@

\subsection{The recursion universe}

@@

  This file specializes the development in \verb|Subrec.v| to the case
  of lists (parametrized by the type \verb|A| of elements).  In
  general, to use our development to get subsidiary recursion over
  some datatype, one will have a similar ``shim'' file.  The file
  defines the signature functor \verb|ListF|, shown in
  Figure~\ref{fig:listf}.  We then define \verb|List| to be
  \verb|Subrec|, with the instantiation of \verb|F| to \verb|ListF A|.
  This type \verb|List| is not to be confused with the type
  \verb|list| of lists in Coq's standard library.  As noted
  previously, our development is meant to be used in extension of
  existing inductive datatypes, not replacing them.  The figure also
  shows constructors \verb|mkNil| and \verb|mkCons| for \verb|List|,
  and types for conversion functions between \verb|List| and
  \verb|list| (see Section~\ref{sec:deriv} for the code).

@@

\subsection{The implementation of \texttt{Subrec} (\texttt{Subrec.v})}

@@

More interesting is the definition of \verb|inn|.

@@

Definition out{R:Set}(fo:FoldT Alg R) : R -> F R :=
  fo F FunF (rollAlg (fun _ _ _ _ d => d)).
\end{verbatim}
\caption{The rest of the interface for \texttt{Subrec}}
\label{fig:subrecb}
\end{figure}

\section{Interface for subsidiary induction}
\label{sec:interfacei}

@@

@@

>>>>

This paper presents a functor-generic derivation of terminating
subsidiary recursion and induction in Coq.  We should emphasize that
this is a derivation of this recursion scheme within the type theory
of Coq.  No axioms or other modifications to Coq of any kind are
required. Based on this derivation, we present several example
functions like \verb|wordsBy|, and prove theorems about them.  A nice
example is a definition of run-length encoding using \verb|span| as a
subsidiary recursion, where we prove that encoding and then decoding
returns the original list.  Our approach applies to the standard
datatypes in the Coq library, and does not require switching libraries
or datatype definitions.

@@

\subsection{The recursion universe}
\label{sec:recu}

@@

  This file specializes the development in \verb|Subrec.v| to the case
  of lists (parametrized by the type \verb|A| of elements).  In
  general, to use our development to get subsidiary recursion over
  some datatype, one will have a similar ``shim'' file.  For space
  reasons, we just give the example of lists.  The file
  defines the signature functor \verb|ListF|, shown in
  Figure~\ref{fig:listf}.  We then define \verb|List| to be
  \verb|Subrec|, with the instantiation of \verb|F| to \verb|ListF A|.
  This type \verb|List| is not to be confused with the type
  \verb|list| of lists in Coq's standard library.  As noted
  previously, our development is meant to be used in extension of
  existing inductive datatypes, not replacing them.  The figure also
  shows constructors \verb|mkNil| and \verb|mkCons| for \verb|List|,
  and types for conversion functions between \verb|List| and
  \verb|list| (see Section~\ref{sec:deriv} for the code).

@@

\subsection{The implementation of \texttt{Subrec} (\texttt{Subrec.v})}
\label{sec:subrecimpl}

@@

More interesting is the definition of \verb|inn|, which is the
critical point where the recursion universe is implemented.  To create
a value of type \verb|Subrec| from data of type \verb|F Subrec|, the
definition of \verb|inn| \verb|roll|s a value of type
\verb|SubrecF Subrec| (we saw this type unfolded at the start of
Section~\ref{sec:subrecimpl}).  This value takes in a carrier
\verb|X|, its fmap function \verb|xmap|, and an algebra \verb|alg|
with that carrier.  Note that the anchor type of this algebra is
\verb|Subrec|.  It will then call \verb|alg| (after \verb|unroll|ing
it) with implementations for the components of the recursion universe
(cf. Section~\ref{sec:recu}, also Figure~\ref{fig:algf}):
\begin{itemize}
\item \verb|Subrec| is passed as the value for the abstract type \verb|R|.
\item the identity function is passed as the value for \verb|reveal : R -> Subrec| (since \verb|R| has
  been instantiated to \verb|Subrec|).
\item The function \verb|fold|, which expects an algebra with anchor type \verb|Subrec|, is passed as
  the fold function of type \verb|FoldT Alg R|.
\item For the \verb|eval : R -> X R| function, we pass \verb|(fold X xmap alg) : Subrec -> X Subrec|.
\item For the subdata structure of type \verb|F R|, we pass \verb|d : F Subrec| (again, since we
  are instantiating \verb|R| with \verb|Subrec|).
\end{itemize}

Finally, Figure~\ref{fig:subrecb} defines \verb|out| as a subsidiary
recursion, given any fold function with its anchor type \verb|R|.  The
code for \verb|out| just folds an algebra over the input of type
\verb|R|, where that algebra simply returns the subdata structure it
is given.  Outside the recusion, this has type \verb|F R|; inside the
recursion it has type \verb|F R'| where \verb|R'| is the abstract type
of the subsidiary recursion (named in the figure just for discussion
here).  So \verb|out| implements the idea that unfolding an abstract
type one step is just a trivial case of subsidiary recursion.

@@

Definition out{R:Set}(fo:FoldT Alg R) : R -> F R :=
  fo F FunF (rollAlg (fun R' _ _ _ d => d)).
\end{verbatim}
\caption{The rest of the interface for \texttt{Subrec}}
\label{fig:subrecb}
\end{figure}

\subsection{Converting between \texttt{list} and \texttt{List} (\texttt{List.v})}

Figure~\ref{fig:listb} gives the definitions (specified in
Figure~\ref{fig:listf} above) for conversion functions between
\texttt{list} (the list type of Coq's standard library) and our
\texttt{List} type for subsidiary recursion on lists.  To convert a
\verb|list| to a \verb|List|, we use \verb|listFold| to recursively
apply the function \verb|inn| (Figure~\ref{fig:subrecb}) throughout a
\verb|list|.  This \verb|listFold| provides a functorial view of the
usual fold-right function on lists.  To convert a \verb|List| to a
\verb|list|, we use our \verb|fold| function
(Figure~\ref{fig:subrecb}) to fold an algebra over the input
\verb|List|.  This algebra does an \verb|fmap| (for \verb|ListF|) with
\verb|eval : R -> list A|.  The resulting value has type
\verb|ListF (list A)|.  We may then apply \verb|listIn| to produce a
\verb|list A|.  This \verb|listIn| function converts from
\verb|ListF (list A)| to \verb|list A| in the obvious way, converting
\verb|Nil| to the empty list, and \verb|Cons| to an application of
\verb|::|.  These functions were not necessary in implementing
examples of subsidiary recursion above, but they will be needed
for reasoning about them.  We turn to this topic next.

\begin{figure}
\begin{verbatim}
Fixpoint listFold(l : list A){X : Set}(alg : ListF X -> X) : X :=
  match l with
    nil => alg Nil
  | cons hd tl => alg (Cons hd (listFold tl alg))
  end.

Definition toList (xs : list A) : List := listFold xs (inn ListF).

Definition listIn(d : ListF (list A)) : list A :=
  match d with
    Nil => []
  | Cons hd tl => hd :: tl
  end.

Definition fromList : List -> list A :=
  fold ListF (Const (list A)) (FunConst (list A))
       (rollAlg (fun R reveal fo eval fr => listIn (fmap eval fr))) .
\end{verbatim}
\caption{Additional functions for the \texttt{List} type (\texttt{Subrec ListF})}
\label{fig:listb}
\end{figure}

\section{Interface for subsidiary induction (\texttt{Subreci.v})}
\label{sec:interfacei}

We have seen how to write subsidiary recursions in Coq.  But can one
reason about these?  To wrap up this paper, we will see an interface
for subsidiary induction in Coq, and example proofs written using this
interface.  Subsidiary induction is written just as the natural
extension of subsidiary recursion, which worked over \verb|Set|s, to
\verb|Subrec|-predicates.  The development is parametrized by a
functor \verb|F| and a functor
\verb|Fi : (Subrec -> Prop) -> (Subrec -> Prop)| over
\verb|Subrec|-indexed propositions (i.e., predicates).  Just as
functors need an \verb|fmap| function, we here need an indexed
version, of type \verb|fmapiT Subrec Fi| (definition elided.)

The central definitions for the type \verb|Subreci : Subrec -> Prop|
are given in Figure~\ref{fig:subreci}.  Where having a value \verb|x|
of \verb|Subrec| entitles us to define subsidiary recursions to
inhabit types \verb|X Subrec|, a value of type \verb|Subreci x| lets
us prove properties of \verb|x| by subsidiary induction.  Briefly:
\verb|kMo| is the kind for \emph{motives}, namely predicates on
\verb|Subrec|~\cite{mcbride00}.  \verb|KAlgi| is the kind for indexed
algebras.  \verb|FoldTi| is the indexed version of \verb|FoldT|: it is
expresses provability of \verb|X C| for \verb|d|, based on an indexed
algebra and a value of type \verb|C d|, where \verb|C| is the anchor
type, now indexed.  \verb|AlgFi| and \verb|Algi| are indexed versions
of the algebras we saw for recursion.  The \verb|eval| function
(Figure~\ref{fig:algf}) has now become an induction hypothesis: given
any \verb|d| where \verb|R d| holds, \verb|ih| proves \verb|X R d|.  A
value of type \verb|R d| is thus a license to induct on \verb|d|.
Finally, the algebra is given a subdata structure indexed by
\verb|d : Subrec|, and must produce a proof of \verb|X R d|.  \verb|Subreci|
is defined as the suitably indexed fixed-point of \verb|SubrecFi|, which
is the natural indexed version of \verb|SubrecF|.

\begin{figure}
\begin{verbatim}
Definition kMo := Subrec -> Prop.
Definition KAlgi := kMo -> (kMo -> kMo) -> Set.
Definition FoldTi(alg : KAlgi)(C : kMo) : kMo :=
  fun d => forall (X : kMo -> kMo) (xmap : fmapiT Subrec X), alg C X -> C d -> X C d.

Definition AlgFi(A: KAlgi)(C : kMo)(X : kMo -> kMo) : Set :=
  forall (R : kMo)
    (reveal : (forall (d : Subrec), R d -> C d))        
    (fo : (forall (d : Subrec), FoldTi A R d))
    (ih : (forall (d : Subrec), R d -> X R d))
    (d : Subrec),
    Fi R d -> X R d.

Definition Algi := MuAlgi Subrec AlgFi.

Definition SubrecFi(C : kMo) : kMo := 
  fun d => forall (X : kMo -> kMo) (xmap : fmapiT Subrec X), Algi C X -> X C d.
Definition Subreci := Mui Subrec SubrecFi.

Definition foldi(i : Subrec) : FoldTi Algi Subreci i.
Definition inni(i : Subrec)(fd : Fi Subreci i) : Subreci i.
\end{verbatim}
\caption{Interface for subsidiary induction}
\label{fig:subreci}
\end{figure}

For lists, we instantiate \verb|Fi| with \verb|ListFi|, shown in
Figure~\ref{fig:listfi}.  This is just the indexed version of
\verb|ListF|.  Given a \verb|list A|, \verb|toListi| returns
a value of type \verb|Listi (toList xs)|.  This can be understood
as saying that for any list (from Coq's standard library),
we can reason by subsidiary induction to prove properties of \verb|toList xs|.
We also introduce an abbreviation \verb|ListFoldTi| for the type
of indexed fold functions over lists.

\begin{figure}
\begin{verbatim}
Definition lkMo := List -> Prop.

Inductive ListFi(R : lkMo) : lkMo :=
  nilFi : ListFi R mkNil
| consFi : forall (h : A)(t : List), R t -> ListFi R (mkCons h t).

Definition Listi := Subreci ListF ListFi.
Definition toListi(xs : list A) : Listi (toList xs) := listFoldi xs Listi inni.
Definition ListFoldTi(R : List -> Prop)(d : List) : Prop :=
  FoldTi ListF (Algi ListF ListFi) R d.
\end{verbatim}
\caption{The indexed version \texttt{ListFi} of \texttt{ListF}}
\label{fig:listfi}
\end{figure}

@@

For proving the main theorem about run-length encoding, we need
several lemmas about \verb|span|, shown in Figure~\ref{fig:spanlem}.
For lack of space, we just state the properties.  The first says that
appending the results of a call to span returns the original list
(module some conversions to \verb|list| from \verb|List|).  The second
uses the inductive type \verb|Forall| from Coq's standard library to
state that all the elements of the prefix returned by \verb|span|
satisfy \verb|p|.  These lemmas are proved using an indexed algebra
where the indexed anchor type is not used (so the carriers are
constant indexed-functor returning the types shown).  But for the
third, \verb|GuardPres|, we are showing that whenever \verb|spanh|
(see Figure~\ref{fig:span}) returns a suffix \verb|r|, that suffix
satisfies the indexed anchor type.  This enables us to invoke an
outer induction hypothesis on this suffix, when using \verb|span|
subsidiarily.

\begin{figure}
\begin{verbatim}
Definition SpanAppendF(p : A -> bool)(xs : List A) : Prop :=
  forall (l : list A)(r : List A) ,
    span p xs = (l,r) ->
    fromList xs = l ++ (fromList r).

Definition spanForallF(p : A -> bool)(xs : List A) : Prop :=
  forall (l : list A)(r : List A),
    span p xs = (l,r) ->
    Forall (fun a => p a = true) l.

Definition GuardPresF(p : A -> bool)(S : List A -> Prop)(xs : List A) : Prop :=
  forall (l : list A)(r : List A),
    spanh p xs = SpanSomeMatch l r ->
    S r.
\end{verbatim}
\caption{Statements of three lemmas about \texttt{span}}
\label{fig:spanlem}
\end{figure}

Using these lemmas, we can write a short proof by subsidiary induction of:

\begin{verbatim}
Theorem RldRle (xs : list A): rld (rle (toList xs)) = xs.  
\end{verbatim}

Finally, as promised, a note on noncanonicity. When proving properties about subsidiary
recursions on \verb|xs : List A|, one should be aware that nothing prevents
the property from being applied to noncanonical \verb|List|s.  For example,
suppose we wish to prove that if all elements of a  list satisfy \verb|p|, then
the suffix returned by \verb|span| is empty.  It is dangerous to phrase
this as ``the suffix equals \verb|mkNil|'', because for a noncanonical input \verb|xs|,
\verb|span| will return that same  noncanonical \verb|xs| as the suffix (and so
it may be a noncanonical empty list, not equal to \verb|mkNil|).  The solution
in this case is to use a function \verb|getNil| (\verb|List.v|) that computes
an empty list from \verb|xs|.  So the statement that one can prove is:

\begin{verbatim}
Definition spanForall2F(p : A -> bool)(xs : List A) : Prop :=
  Forall (fun a => p a = true) (fromList xs) ->
  span p xs = (fromList xs, getNil xs).
\end{verbatim}

@@

\section{Conclusion}
